Oak resilience to drought and land-use show how the ecological and geographical rear-edges do not necessarily meet today

### 2019-12-07 man/ecosystems/ms_ecosystems.Rmd
update english review

### Short title: *Relict oak resilience at the rear-edge*

### Authors [^\*^] 
A.J. Pérez-Luque^1,2,a^; G. Gea-Izquierdo^3,b^ and Zamora, R.^1,2,c^

@@

[^\*^]: Author Contributions. AJPL, GGI and RZ conceived the study, conducted field work and collected the data. AJPL and GGI conducted lab work. AJPL analyzed data and led the writing of the paper. GGI and RZ contributed in the writing process. All authors contributed to the drafts and gave final approval for publication.

### Manuscript highlights
* *Quercus pyrenaica* rear-edge forests showed high resilience at tree and stand-levels
* The growth response to climate and resilience followed a water-stress gradient
* Trees and stand expressed high sensitivity to drought and land-use legacies

@@

* Global change challenges ecosystems in dry locations transformed by intensive human use. Forest resilience to drought of relict Mediterranean *Quercus pyrenaica* Willd. populations in the Southern Iberian Peninsula was analyzed in relation to historical records of land-use using dendroecological growth of adult trees and greenness (EVI) as proxies for secondary and primary growth.

* The growth trends reflected a strong influence of land-use legacies (*e.g.* firewood extraction) in the current forest structure. Trees were highly sensitive to moisture availability but both primary and secondary growth expressed high resilience to drought events in the short- and the long-term. The response to climate and resilience followed a water-stress gradient. Primary and secondary growth were generally less resilient to drought in mesic than in xeric sites, whereas a positive growth trend matching the warming climate since the late 1970s was particularly evident in mesic (*i.e.* colder and more humid) high-elevation stands, but absent in the most xeric site.

* Adult oak trees showed high resilience to drought despite being considered relict and at its rear-edge. The trees exhibited a variable response to the changing climate along a very short climatic gradient and showed that the ecological and geographical rear-edges do not need to meet today in forests with strong land-uses legacies. Forest resilience including that of regeneration needs to be monitored at xeric of the species under future expected climatic scenarios bearing in mind that ecological boundaries for species distribution can be narrow and that land-use legacies can play a key role in stand dynamics.

### Keywords 
extreme drought, resilience, rear-edge, *Quercus pyrenaica*, tree-growth, dendrocronology, remote-sensing

## Introduction
The response of species to changing environments is likely to be largely determined by population responses at range margins [@Hampe2005]. Peripheral populations are usually considered more vulnerable compared with populations occurring at the center of a species' range (*i.e.* centre-periphery hypothesis) [@Sagarin2002; @Pironon2016]. It has been assumed that rear-edge populations have a concordance between geographical and ecological marginality, with lower performance and thus higher risk of extinction than those populations at the core of the species' range [@VilaCabrera2019; @Pironon2016]. Nonetheless, recent reviews found species- and population- specific responses highlight that this hypothesis not always fulfill [@Sexton2009; @Abeli2014]. In this respect, to fully understand changes in distribution and abundance of species as a consequence of global change we need to know in which environmental conditions geographical marginality does not correspond with ecological marginality [@VilaCabreraJump2019GreaterGrowth].

Climate change is expected to cause important changes in the distribution and abundance of plant communities, and there are already obvious signs of processes associated with increased intensity and duration of droughts [@Allen2010]. In fact, there has been an increase in the frequency and severity of drought events in the last decades, along with a trend towards drier summers particularly for southern Europe [@VicenteSerrano2014; @Stagge2017]. This is especially critical, since climate change projections for the Mediterranean area forecast that climate extreme events will become more frequent and severe [@Spinoni2017a]. Under this climatic change context, it is reasonable to expect that population loss and range retractions in boreal, temperate and Mediterranean species should be seen in the lowest latitudes, altitudes and drought-prone areas of a species' distribution, *i.e.* the rear edge [@Hampe2005]. In a global-change context, the impacts of drought on vegetation need to be evaluated considering the interactions with other drivers of change [@DoblasMiranda2017]. This is especially relevant for areas with a long history of landscape modification, such us the Mediterranean region, where land-use change plays a significant role in current ecosystem ecology [@Navarro2013], considering also that the interactions of drought and land-use change are crucial for Mediterranean forests [@DoblasMiranda2017].

The impacts of drought are also especially relevant for populations considered to be located in the rear-edge (either latitudinal or altitudinal along climatic gradients) of their distribution, where climatic conditions mark the species-limits and the populations are likely to be more sensitive to small variations on climate and the microtopography [@Hampe2005]. The response of rear-edge populations to variations on climate could be useful to estimate the adaptation of species to predicted climatic changes [@Jump2010]. Yet it is crucial to accurately characterize the ecological rear-edge of species. This can be particularly misleading in ecosystems with strong land-use legacies because it is established from current species distribution [@VilaCabrera2019]. Furthermore, it becomes even more important for relict populations driven by historical land-uses if they harbour high levels of intraspecific genetic diversity [@Valbuena2013].

Understanding the vegetation response to drought requires of integrative and multidisciplinary strategies to assess the impacts of disturbances on forest at several scales [@Hartman2018]. The assessment of resilience to climate disturbances like droughts provides critical information about the capacity of the forests to maintain their structure and provide valuable ecosystem services. The resilience to drought of species is better assessed when several approaches like dendrocronology and remote sensing are combined simultaneously. Dendroecological estimates of growth (*i.e.* tree-ring width) are a widely used proxy to study tree vitality and the analysis of annual tree-ring widths can be used to study changes in growth as response to drought at the individiual tree-scale [@Fritts1976]. Remote sensing can be used for studying droughts impacts on ecosystems, hence the stand level [*e.g.* @Zhang2013]. The combination of both approaches (remote sensing and dendroecology), has been used to assess the effects of droughts on vegetation along ecological gradients [*e.g.* @VicenteSerrano2013; @Coulthard2017], and also for the analysis of growth resilience to drought on several tree species [*e.g.* @Gazol2018; @PenaGallardo2018]. Nonetheless, it is crucial to know if the responses at the tree level differ from responses at the ecosystem level and characterize the spatial variability of this response in rear-edge populations.

In this study we used remote sensing information and dendroecological methods to evaluate the impact of drought in both canopy greenness (as a proxy to primary growth) and tree-radial growth (as a proxy to secondary growth) of *Quercus pyrenaica* Willd. (*melojo oak*) on southern relict forests at the rear-edge of the species distribution and where species performance is considered to be severely threatened by climate change. We also assessed the resilience of these forests both to several extreme drought episodes and to climate change (*i.e.* warming) in the long-term in the last decades. Our main hypothesis is that these relict populations driven by historical land-use are particularly vulnerable to climate change at their present climatic (either altitudinal or latitudinal) rear-edge, and will show low resilience after extreme drought particularly at the dry edge along climatic gradients at the small scale within the rear-edge. To address this hypothesis our specific aims were: (*i*) To quantify how recent extreme drought events influenced primary and secondary growth of *Q. pyrenaica* forests in its rear edge; (*ii*) to analyze the long-term resilience of these forests to extreme drought events using time-series of radial growth; (*iii*) and to explore differences in the resilience metrics between populations located in contrasting ecological conditions (*i.e.* xeric *vs.* mesic) within the rear edge to detect vulnerability to climate change along climatic gradients at the small spatial scale. To achieve goal (*iii*) and characterize variability in the forest response to drought within the current geographical rear-edge we assessed whether the effect of aspect and environmental conditions expressed in northern and southern populations of *melojo* oak forests differ in their resistance, resilience and recovery to extreme drought events.

@@

## Materials and methods 
### Tree species and study site 
*Quercus pyrenaica* forests extend through south-western France and the Iberian Peninsula reaching its southern limit in mountains areas of northern Morocco [@Franco1990]. In the Iberian Peninsula these forests live on siliceous soils under meso-supramediterranean and mesotemperate areas and subhumid, humid and hyperhumid ombroclimate. *Q. pyrenaica* is a deciduous species that requires over 650 mm of annual precipitation and some summer precipitation. It is a submediterranean species, hence with lower drought tolerance than genuine evergreen Mediterranean taxa [@delRio2007].

The forests of this species reach their southernmost European limit in mountains from Southern Andalusia like Sierra Nevada (37°N, 3°W), a high-mountain range with elevations up to 3482 m *a.s.l.*. The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought but with marked variability set by altitude. Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species [@Olalde2002]. There are eight melojo oak patches (2400 ha) identified in this mountain range (Figure 1), ranging between 1100 and 2000 m *a.s.l.* and often associated to major river valleys. Today, *Q. pyrenaica* woodlands in this mountain region represent a rear edge of their habitat distribution [@Hampe2005]. They are the richest vascular-plant forest formation of Sierra Nevada, sheltering several endemic and endangered plant species [@Loriteetal2008ABG]. These relict forests undergone intensive anthropic use in history [@CamachoOlmedo2002]. Furthermore, conservation status of this species for southern Spain is considered "Vulnerable" and it is expected to suffer the impact of climate change, reducing its suitable habitats in the near future [@GeaIzquierdo2013].

### Drought episodes
The Iberian Peninsula suffered several extreme drought episodes in the last three decades (*e.g*. 1994-1995, 1999, 2005, 2012) [@VicenteSerrano2014]. The 2005 and 2012 drought events have been documented among the worst in the last decades for the southern Iberian Peninsula [@Pascoa2017] and they were characterized as extreme drought in our climatic data (Figure S1; Table S3). We focused on these two drought events because they were included in the period where there is availability of high-spatial resolution of remote sensing information (MODIS started on 2000; see below). Nevertheless, for radial growth-time series, a greater number of older drought events were also analyzed to contextualize results obtained in 2005 and 2012 and to analyze forest resilience to drought in a longer term (see Table S3). The identification of a drought event were done using the Standardized Precipitation-Evapotranspiration Index (SPEI) [@VicenteSerrano2010] following similar procedure as proposed by @Spinoni2015. We used SPEI-12 months scale for all 0.5º grid cells covering Sierra Nevada obtained from Global SPEI Database (http://spei.csic.es/database.html). A severe drought event starts when SPEI falls below the threshold of -1.28 [@Spinoni2017a; @Pascoa2017]. A drought event is considered only when SPEI values are below that threshold for at least two consecutive months. For each drought event we computed: the *duration*, *i.e.* number of consecutive months with the SPEI lower than a certain threshold; the *severity* as the sum of the SPEI values (absolute values) during the duration of the drought event; the *intensity* and the *Lowest SPEI* refer to the mean and lowest value of SPEI respectively during the drought event.

### Greenness data to assess ecosystem resilience
Vegetation greenness of *Q. pyrenaica* was characterized by means of the *Enhanced Vegetation Index* (EVI), derived from MOD13Q1 product of the MODIS sensor. EVI data consists of 16-day maximum value composite images (23 per year) of the EVI value with a spatial resolution of 250 m x 250 m. MODIS EVI data were obtained for the period 2000 - 2016. We selected the pixels covering the distribution of *Q. pyrenaica* forests in Sierra Nevada (*n* = 928 pixels). Values affected by high content of aerosols, clouds, snow and shadows, were filtered out following recommendations for mountain regions [@Reyes2015].

Annual mean EVI ($EVI_{mean}$) as a surrogate of mean annual primary production was computed for each pixel for the period 2000 - 2016. The EVI standardized anomaly ($EVI_{sa}$) was computed pixel-by-pixel, to minimize bias in the evaluation of anomalies and provide more information about their magnitude [@Samanta2012]. For each pixel, an annual EVI value was obtained by averaging EVI valid values. Then, the standardized anomaly was computed as: $\mathrm{EVI_{sa,\mathit{i}}}= (\mathrm{EVI_{mean,\mathit{i}}-EVI_{mean,ref}})\Big/\sigma_{\mathrm{ref}}$, 
where $\mathrm{EVI_{sa,\mathit{i}}}$ is the EVI standardized anomaly for year $i$;  $\mathrm{EVI_{mean,\mathit{i}}}$ the annual mean value of EVI for year $i$; $\mathrm{EVI_{mean,ref}}$ the average of the annual EVI values for the period of reference 2000-2016 (all except year $i$); and $\sigma_{\mathrm{ref}}$ the standard deviation for the reference period. Each pixel was categorized according the EVI standardized anomalies as "greening" ($EVI_{sa} > 1$), "browning" ($EVI_{sa} <- 1$) or "no-changes" ($-1 > EVI_{sa} > 1$)[@Samanta2012].

Rather than other vegetation indices like the NDVI, $EVI_{mean}$ was chosen since it is highly stable under the use of any filter [@Reyes2015] and because it showed high significant correlations with annual ($r$= 0.81) and seasonal EVI values ($r_{spring}$= 0.76 and $r_{summer}$= 0.88). For exploratory purposes, temporal trends of $EVI_{mean}$ were explored at a pixel scale, using Mann–Kendall nonparametric test.

### Field sampling and dendrochronological methods to assess tree individual resilience
Tree sampling was carried out during autumn 2016. Trees were sampled at two locations in contrasting N-S slopes of Sierra Nevada: San Juan (SJ), a xeric site located at the northern aspect; and Cáñar (CA), a more humid site located at the southern aspect (Figure 1; Table 1). For the southern site two elevations were sampled: CA-Low (around 1700 m) and CA-High (around 1860 m), which constitute the current low altitudinal limit (CA-Low) and the tree-line (CA-High), respectively in the sampled aspect. Despite the proximity of these two elevations (less than 200 m difference in altitude) the stands greatly differ in their structure and characteristics (Table 1). The three sampling sites followed a moisture gradient: SJ < CA-Low < Ca-High (Table 1). All the sites were oak monospecific and representative of the population clusters identified for the species in this mountain range [@PerezLuque2015]. In each site between 15 and 20 trees from either the single dominant-codominant layer in CA or the open canopy in SJ were randomly selected. Two cores of 5 mm of diameter were taken from each tree at breast heigth (1.3 m) using an increment borer. Diameter at breast height (DBH) and total height were measured for each tree. In addition, stand competition affecting target trees was assessed by recording distance, azimuth, DBH, species and total height of all neighboring living trees with DBH > 7.5 cm within a circular plot of 10 m radius. Several competition indices were calculated: the distance independent indices *density* ($\mathrm{trees \cdot ha^{-1}}$), and *basal area* (BA, $\mathrm{m^{2} \cdot ha^{-1}}$); and the distance dependent index size ratio proportional to distance as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$ [see @GeaIzquierdo2009 for more details]. Differences between sites for height, DBH and competition indices were analysed using non-parametric Kruskal-Wallis rank sum tests. When significant differences were observed, we run multiple comparisons using the Dunn's-test with Bonferroni adjustment to correct for significance.

Tree cores were air dried, glued onto wooden mounts and sanded. Annual radial growth (ring width, RW) was measured with a measuring device coupled to a stereomicroscope, with an accuracy of 0.001 mm. Individual ring series were first visually and statistically cross-dated with TSAP software (Rinntech, Heidelberg, Germany), using the statistics Gleichläufigkeit (GLK), t-value and the crossdating index (CDI). Cross-dating validation was finally verified using COFECHA [@Holmes1983].

The growth trends were analyzed at different time scales. To study the response of growth to the inter-annual variability of climate (short-term response), pre-whitened residual chronologies (RWI) were used. They were calculated from ratios between raw growth measurements and individual cubic splines with a 50 % frequency cutoff at 30 years [@Fritts1976]. Tree-ring width series were standardized and detrended using `dplR` [@Bunn2010]. Mean residual site chronologies were obtained computing the biweight robust mean of all prewhitened growth indices for the trees of the same site [@Fritts1976]. The statistical quality of each chronology was checked via the expressed population signal (EPS). A threshold value of EPS > 0.85 was used to determine the cutoff year of the time span that could be considered reliable.

The long-term growth response was analyzed using basal area increment (hereafter BAI, $\mathrm{cm^2 \cdot year^{-1}}$). Theoretically, BAI represents a more accurate indicator of growth than ring-width, since it removes variation in growth attributable to increasing stem circumference after 30-40 years of juvenile increasing growth [@Biondi2008]. Raw ring-widths and measured DBH were used to compute BAI [@Piovesan2008] using the following equation: $\mathrm{BAI} = \pi(r^2_{t}-r^2_{t-1})$ where $r$ is the radius of the tree and $t$ is the year of tree-ring formation. For each individual tree, a mean BAI series was calculated. Then, mean site BAI chronologies were obtained by averaging individual tree BAI time series.

### Climate and growth 
Climate data were obtained from the European Daily High-Resolution Observational Gridded Dataset (E-OBS v16) [@Haylock2008]. Monthly precipitation and minimum and maximum temperatures were obtained at a 0.25 x 0.25 º resolution for the 1950-2016 period. We selected grid cells covering each sampled sites. SPEI index with a temporal scale of 6 months was used to characterize the drought conditions for the period 1961-2014. Relationships between RWI and BAI site chronologies and climatic variables were assessed using bootstrapped Pearson's correlations estimated using `treeclim` [@Zang2015].

>>>>

### Short title: *Relict oak resilience at the rear edge*

### Authors [^\*^] 
A.J. Pérez-Luque^1,2,a^; Gea-Izquierdo, G.^3,b^ and Zamora, R.^1,2,c^

@@

[^\*^]: Author Contributions. AJPL, GGI and RZ conceived of the study, conducted field work, and collected the data. AJPL and GGI performed the lab work. AJPL analyzed data and led the writing of the paper. GGI and RZ contributed in the writing process. All authors contributed to the drafts and gave final approval for publication.

### Manuscript highlights
* *Quercus pyrenaica* rear-edge forests showed high resilience at tree and stand levels
* Resilience and growth response to climate followed a water-stress gradient
* Trees and stand expressed high sensitivity to drought and land-use legacies

@@

* Global change challenges ecosystems in dry locations transformed by intensive human use. Forest resilience to drought of relict Mediterranean *Quercus pyrenaica* Willd. populations in the southern Iberian Peninsula was analyzed in relation to historical records of land use employing dendroecological growth of adult trees and greenness (EVI) as proxies for secondary and primary growth.

* The growth trends reflected a strong influence of land-use legacies (*e.g.* firewood removal) in the current forest structure. Trees were highly sensitive to moisture availability but both primary and secondary growth expressed high resilience to drought events over the short and the long term. Resilience and response to climate followed a water-stress gradient. Primary and secondary growth were generally less resilient to drought in mesic than in xeric sites, whereas a positive growth trend matching the warming climate since the late 1970s was particularly evident in mesic (*i.e.* colder and wetter) high-elevation stands, but absent in the most xeric (*i.e.* warmer and drier) site.

* Adult oak trees showed high resilience to drought despite being considered relict and at its rear edge. The trees varied in response to the changing climate along a narrow climatic gradient and showed that the ecological and geographical rear edges do not need to meet today in forests with strong land use legacies. Forest resilience including that of regeneration needs to be monitored at xeric of the species under future expected climatic scenarios given that ecological boundaries for species distribution can be narrow and that land-use legacies can play a key role in stand dynamics.

### Keywords 
extreme drought, resilience, rear edge, *Quercus pyrenaica*, tree growth, dendrochronology, remote sensing

## Introduction
The response of species to changing environments is likely to be determined largely by population responses at range margins [@Hampe2005]. Peripheral populations are usually considered more vulnerable compared with populations occurring at the center of a species' range [*i.e.* center-periphery hypothesis; @Sagarin2002; @Pironon2016]. It has been assumed that rear-edge populations have a concordance between geographical and ecological marginality, with lower performance, higher vulnerability, and thus higher risk of extinction than do populations at the core of the species' range [@VilaCabrera2019; @Pironon2016]. Nonetheless, recent reviews report that species- and population-specific responses do not always support this hypothesis [@Sexton2009; @Abeli2014]. In this respect, to fully understand changes in distribution and abundance of species as a consequence of global-change, we need to determine 
under what environmental conditions geographical marginality does not correspond with ecological marginality [@VilaCabreraJump2019GreaterGrowth].

Climate change is expected to cause major shifts in the distribution and abundance of plant communities, and there are already signs of processes associated with increased intensity and duration of droughts [@Allen2010]. In fact, the frequency and severity of drought events
have increased in recent decades, along with a trend towards drier summers, particularly for Southern Europe [@VicenteSerrano2014; @Stagge2017]. This is especially critical, since climate-change projections for the Mediterranean area forecast more frequent and severe extreme climate events [@Spinoni2017a]. In this climatic-change context, population loss and range retractions are expected in boreal, temperate, and Mediterranean species at the lowest latitudes, elevations, and drought-prone areas of a species' distribution, *i.e.* the rear edge [@Hampe2005]. In a global-change context, the impact of drought on vegetation needs to be evaluated regarding the interactions with other drivers of change [@DoblasMiranda2017]. This is especially relevant for areas with a long history of landscape alteration, such as the Mediterranean region, where land-use change plays a major role in current ecosystem ecology [@Navarro2013], given that the interactions of drought and land-use change are crucial for Mediterranean forests [@DoblasMiranda2017].

The effects of drought are also especially relevant for populations considered to be located along the rear edge (either latitudinal or altitudinal along climatic gradients) of their distribution, where climatic conditions mark the species limits, and the populations are likely to be more sensitive to minor climatic and microtopographic variations [@Hampe2005]. The response of rear-edge populations to climate variations could be useful in estimating the adaptation of species to predicted climatic changes [@Jump2010]. Yet it is crucial to accurately characterize the ecological rear-edge of species. This can be particularly misleading in ecosystems with strong land-use legacies because it is established from current species distribution [@VilaCabrera2019]. Furthermore, this characterization becomes even more critical for relict populations driven by historical land uses when they harbor high levels of intraspecific genetic diversity [@Valbuena2013].

Understanding the vegetation response to drought requires integrative and multidisciplinary strategies to assess the disturbances of forest at several scales [@Hartman2018]. The assessment of resilience to climate disturbances such as droughts provides critical information concerning the capacity of the forests to maintain their structure and render valuable ecosystem services. The drought resilience of species is better assessed when several approaches, such as dendrochronology and remote sensing, are combined. Dendroecological estimates of growth (*i.e.* tree-ring width) are commonly used proxies to characterize tree vitality, and annual tree-ring widths can be used to study changes in growth as a response to drought at the individiual-tree scale [@Fritts1976]. Remote sensing can be used to study the impact of drought on ecosystems and hence the stand level [*e.g.* @Zhang2013]. The combination of both approaches (remote sensing and dendroecology) has been used to assess the effects of droughts on vegetation along ecological gradients [*e.g.* @VicenteSerrano2013; @Coulthard2017], and also to evaluate growth resilience to drought in several tree species [*e.g.* @Gazol2018; @PenaGallardo2018]. Nonetheless, it is crucial to ascertain wether the responses at the tree level differ from those at the ecosystem level and characterize the spatial variability of this response in rear-edge populations.

In the present study, we used remote-sensing information and dendroecological methods to evaluate the impact of drought in both canopy greenness (as a proxy to primary growth) and radial tree growth (as a proxy for secondary growth) of *Quercus pyrenaica* Willd. (*melojo oak*) in southern relict forests at the rear-edge of the species distribution and where species performance is considered to be severely threatened by climate change. We also assessed the resilience of these forests both to several extreme drought episodes and to climate change (*i.e.* warming) over the long term (in the last few decades). Our main hypothesis is that these relict populations driven by historical land use are vulnerable to climate change at their present climatic (either elevational or latitudinal) rear edge, and hence will show low resilience to extreme drought along climatic gradients at a small scale within the rear edge. To test this hypothesis, we: (*i*) quantified how recent extreme drought events influenced primary and secondary growth of *Q. pyrenaica* forests at their  present geographical rear edge; (*ii*) analyzed the long-term resilience of these forests to extreme drought events, using time-series of radial growth; (*iii*) and examined differences in the resilience metrics between populations located in contrasting ecological conditions (*i.e.* xeric *vs.* mesic) along environmental gradients within the rear edge in order to detect vulnerability to climate change at the small spatial scale. For the latter task, we characterized the variability in the forest response to drought within the current geographical rear edge and assessed whether the effect of aspect and environmental conditions expressed in northern and southern populations of *melojo* oak forests differ in their resistance, resilience and recovery to extreme drought events.

@@

## Materials and methods 
### Tree species and study site 
*Quercus pyrenaica* forests extend throughout south-western France and the Iberian Peninsula reaching their southern limit in mountain areas of northern Morocco [@Franco1990]. In the Iberian Peninsula, these forests occupy siliceous soils under meso-supramediterranean and mesotemperate areas and subhumid, humid, and hyperhumid ombroclimate. *Q. pyrenaica* is a deciduous species that requires over 650 mm of annual precipitation and some summer precipitation. As a submediterranean species, it has lower drought tolerance than evergreen Mediterranean taxa [@delRio2007].

The forests of this species reach their southernmost European limit in Andalusian mountains such as Sierra Nevada (37°N, 3°W), a high-mountain range with elevations of up to 3482 m *a.s.l.*. The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought but with marked variability according to elevation. Sierra Nevada is considered a glacial refuge for deciduous *Quercus* species [@Olalde2002]. Eight melojo oak patches (2400 ha) have been identified in this mountain range (Figure 1), from 1100 to 2000 m *a.s.l.* and often associated with major river valleys. Today, *Q. pyrenaica* woodlands in this mountain region represent a rear edge of their habitat distribution [@Hampe2005]. They are the richest forest formation of Sierra Nevada, sheltering several endemic and endangered plant species [@Loriteetal2008ABG]. These relict forests have undergone intensive human use throughout history [@CamachoOlmedo2002]. Furthermore, the conservation status of this species for southern Spain is considered "Vulnerable" and it is expected to suffer from climate change, reducing its suitable habitats in the near future [@GeaIzquierdo2013].

### Climatic data and drought episodes
Climate data were obtained from the European Daily High-Resolution Observational Gridded Dataset (E-OBS v16) [@Haylock2008]. Monthly precipitation and minimum and maximum temperatures had a 0.25 x 0.25 º resolution for the 1950-2016 period. Grid cells were select to cover each sampled site. The SPEI (Standardized Precipitation-Evapotranspiration Index) [@VicenteSerrano2010] index with a temporal scale of 6 months was used to characterize the drought conditions for the period 1961-2014.

The Iberian Peninsula underwent several extreme drought episodes in the last three decades (*e.g*. 1995, 1999, 2005, 2012) [@VicenteSerrano2014]. The 2005 and 2012 drought events have been documented as being among the worst in recent decades for the southern Iberian Peninsula [@Pascoa2017], and they were characterized as extreme drought in our climatic data (Figure S1; Table S3). We focused on these two drought events because they were included in the period having remote-sensing information of high spatial resolution (MODIS started on 2000; see below). Nevertheless, for radial growth-time series, a greater number of older drought events were also analyzed to contextualize the results for 2005 and 2012 and to evaluate forest resilience to drought over a longer term (see Table S3). A drought event was identified using the SPEI 12-months scale, following a procedure similar to the one proposed by @Spinoni2015. We used 0.5º grid cells covering Sierra Nevada taken from the Global SPEI Database (http://spei.csic.es/database.html). A severe drought event starts when SPEI falls below the threshold of -1.28 [@Spinoni2017a; @Pascoa2017]. A drought event is considered only when SPEI values fall below that threshold for at least two consecutive months. For each drought event, we computed: the *duration*, *i.e.* number of consecutive months with the SPEI lower than a certain threshold; the *severity*, *i.e.* the sum of the SPEI values (absolute values) during the drought event; the *intensity* and the *Lowest SPEI* refer to the mean and lowest value of SPEI respectively during the drought event.

### Greenness data to assess ecosystem resilience
Vegetation greenness of *Q. pyrenaica* was characterized by means of the *Enhanced Vegetation Index* (EVI), derived from MOD13Q1 product of the MODIS sensor. EVI data consists of 16-day maximum value composite images (23 per year) of the EVI value with a spatial resolution of 250 m x 250 m. MODIS EVI data were compiled for the period 2000 - 2016. We selected the pixels covering the distribution of *Q. pyrenaica* forests in Sierra Nevada (*n* = 928 pixels). Any values affected by clouds, snow, shadows or high content of aerosols, were filtered out following recommendations for mountain regions [@Reyes2015].

The mean Annual EVI ($EVI_{mean}$) as a surrogate of mean annual primary production was computed for each pixel for the period 2000 - 2016. The EVI standardized anomaly ($EVI_{sa}$) was computed pixel-by-pixel, in order to minimize bias in the evaluation of anomalies and to provide more information concerning their magnitude [@Samanta2012]. For each pixel, an annual EVI value was calculated by averaging EVI valid values. Then, the standardized anomaly was computed as: $\mathrm{EVI_{sa,\mathit{i}}}= (\mathrm{EVI_{mean,\mathit{i}}-EVI_{mean,ref}})\Big/\sigma_{\mathrm{ref}}$, 
where $\mathrm{EVI_{sa,\mathit{i}}}$ is the EVI standardized anomaly for year $i$;  $\mathrm{EVI_{mean,\mathit{i}}}$ the annual mean value of EVI for year $i$; $\mathrm{EVI_{mean,ref}}$ the average of the annual EVI values for the period of reference 2000-2016 (all except year $i$); and $\sigma_{\mathrm{ref}}$ the standard deviation for the reference period. Each pixel was categorized according the EVI standardized anomalies as "greening" ($EVI_{sa} > 1$), "browning" ($EVI_{sa} <- 1$) or "no-changes" ($-1 > EVI_{sa} > 1$)[@Samanta2012].

Rather than other vegetation indices such as the NDVI, $EVI_{mean}$ was chosen because it is highly stable under the use of any filter [@Reyes2015] and because it showed highly significant correlations with annual ($r$= 0.81) and seasonal EVI values ($r_{spring}$= 0.76 and $r_{summer}$= 0.88). For exploratory purposes, temporal trends of $EVI_{mean}$ were examined at the pixel scale, using the Mann–Kendall nonparametric test.

### Field sampling and dendrochronological methods to assess individual tree resilience
Trees were sampled during autumn 2016 at two locations in contrasting N-S slopes of Sierra Nevada: San Juan (SJ), a xeric site located at the northern aspect; and Cáñar (CA), a wetter site located at the southern aspect (Figure 1; Table 1). For the southern site, two elevations were sampled: CA-Low (around 1700 m) and CA-High (around 1860 m), constituting the current low-elevational limit (CA-Low) and the tree-line (CA-High), respectively, in the sampled aspect. Despite the proximity of these two elevations (less than a 200-m difference) the stands differ markedly in their structure and characteristics (Table 1). The three sampling sites followed a moisture gradient: SJ < CA-Low < Ca-High (Table 1). All the sites were oak monospecific and representative of the population clusters identified for the species in this mountain range [@PerezLuque2015]. At each site, between 15 and 20 trees from either the single dominant-codominant layer in CA or the open canopy in SJ were randomly sampled. Two cores of 5 mm of diameter were taken from each tree at breast height (1.3 m) using an increment borer. Diameter at breast height (DBH) and total height were measured for each tree. In addition, stand competition affecting target trees was assessed by recording distance, azimuth, DBH, species, and total height of all neighboring living trees with DBH > 7.5 cm within a circular plot with a 10-m radius. Several competition indices were calculated: the distance independent indices *density* ($\mathrm{trees \cdot ha^{-1}}$), and *basal area* (BA, $\mathrm{m^{2} \cdot ha^{-1}}$); and the distance dependent index size ratio proportional to distance as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$ [see @GeaIzquierdo2009 for more details]. Differences between sites for height, DBH and competition indices were analyzed using non-parametric Kruskal-Wallis rank sum tests. When significant differences were detected, multiple comparisons were run using the Dunn's-test with Bonferroni adjustment to correct for significance.

Tree cores were air dried, glued onto wooden mounts, and sanded. Annual radial growth (ring width, RW) was determined with a measuring device coupled to a stereomicroscope, for an accuracy of 0.001 mm. Individual ring series were first visually and statistically cross-dated with TSAP software (Rinntech, Heidelberg, Germany), using the statistics Gleichläufigkeit (GLK), t-value and the crossdating index (CDI). Cross-dating validation was finally verified using COFECHA [@Holmes1983].

The growth trends were analyzed at different time scales. To study the growth response to the inter-annual variability of climate (short-term response), pre-whitened residual chronologies (RWI) were used. These were calculated from ratios between raw growth measurements and individual cubic splines with a 50% frequency cutoff at 30 years [@Fritts1976]. Tree-ring width series were standardized and detrended using `dplR` [@Bunn2010]. Mean residual site chronologies were established by computing the biweight robust mean of all prewhitened growth indices for the trees of the same site [@Fritts1976]. The statistical quality of each chronology was checked via the expressed population signal (EPS). A threshold value of EPS > 0.85 was used to determine the cutoff year of the time span that could be considered reliable.

The long-term growth response was analyzed using basal area increment (hereafter BAI, $\mathrm{cm^2 \cdot year^{-1}}$). In theory, BAI represents a more accurate indicator of growth than ring width, since it removes variation in growth attributable to increasing stem circumference after 30-40 years of juvenile growth [@Biondi2008]. Raw ring widths and measured DBH were used to compute BAI [@Piovesan2008] with the following equation: $\mathrm{BAI} = \pi(r^2_{t}-r^2_{t-1})$ where $r$ is the radius of the tree and $t$ is the year of tree-ring formation. For each individual tree, a mean BAI series was calculated. Then, mean site BAI chronologies were determined by averaging individual tree BAI time series.

### Disturbance analyses 
Disturbance chronologies were built using tree-ring width to identify abrupt and sustained increases (release events from competition) or decreases (supressions) in radial growth [@Nowacki1997] as indirect estimates of possible disturbance events (*e.g.* logging, drought-induced neighbor mortality) in the past. Growth changes (GC) were calculated for the individual tree-ring series using a 10-year running window as either positive (PGC) or negative (NGC) growth changes: $\mathrm{\% GC} = \left[(M1 - M2)\Big/ M2\right] \times 100$, where $M1$ is the preceding 10-year median and $M2$ is the subsequent 10-year median [@Rubino2004].
```{r, eval=FALSE, echo=FALSE}
# Medians were used since they are a more robust estimator of central tendency than means [@Rubino2004].
```
Site-disturbance chronologies were constructed by annually averaging the individual disturbance series. To separate growth peaks caused by disturbance events and expressing stand-wise disturbances from those by climate, we considered a threshold of 50% of GC and more than 50% of the individual trees displaying the same growth changes [*e.g.* @GeaIzquierdo2014]. In addition, the history of the forest and management of our sampling sites was inferred from a detailed analysis of historical land-use changes. For this, existing historical documents were exhaustively reviewed to compile information on socio-economical activities affecting the forests being studied (Table S4). We reviewed several documentary sources: historical documents and maps; detailed mining reports; official information on recent wildfires events and forest-management practices; livestock farming; traditional irrigation channels; and studies concerning the socioeconomic dynamics of forests on Sierra Nevada at different scales (see Table S4 for references).
